Learning Discrete Bayesian Networks from Continuous Data

### 2017-05-11 jair_Learning_Discrete_Bayesian_Networks_from_Continuous_Data.tex
mjk edits

Learning Bayesian networks from raw data can help interpret the variable interrelationships and structure without relying on field-specific experts.
While real data often contains a mixture of discrete and continuous-valued variables, many Bayesian network structure learning algorithms assume all random variables are discrete.
Thus, continuous variables are often discretized when learning a Bayesian network.
However, the choice of discretization policy has significant impact on the accuracy, speed, and interpretability of the resulting models.
This paper introduces a principled Bayesian discretization method for continuous variables in Bayesian networks with quadratic complexity instead of the cubic complexity of other standard techniques.
Empirical demonstrations show that the proposed method is superior to the established minimum description length algorithm.
In addition, this paper shows how to incorporate existing methods into the structure learning process to discretize all continuous variables and simultaneously learn Bayesian network structures.

@@

Bayesian networks \citep{Pearl_1988,PGM_2009} are often used to model uncertainty and causality, with applications ranging from decision-making systems \citep{kochenderfer2012next} to medical diagnosis \citep{Lustgarten_2011}. Learning Bayesian networks from raw data can help users to discover knowledge without field-specific experts, such as causality and probabilistic relation between variables. It is common to assume that the random variables in a Bayesian network are discrete, since many Bayesian network learning algorithms are unable to efficiently handle continuous variables. In addition, many of the commonly used Bayesian network software packages, such as Netica \citep{netica1992}, SMILearn \citep{druzdzel1999smile}, and bnlearn \citep{bnlearn2010}, are geared towards discrete variables. However, many applications require the use of continuous data, such as position and velocity in dynamic systems \citep{kochenderfer2010airspace}.

@@

Let $\cX$ be a continuous variable and let $\cx$ be a specific instance of $\cX$.
A discretization policy $\Lambda_{\cX} = \anglebrackets{e_1 < e_2 < \ldots < e_{k-1}}$ for $\cX$ is a mapping from $\mathbb{R}$ to $\curly{1,2,3,\ldots,k}$ such that
\begin{equation}
  \Lambda_{\cX}(\cx) = \begin{cases}
    1, & \text{if $x<e_1$}\\
    i, & \text{if $e_{i-1} \leq x < e_i$}\\
    k, & \text{otherwise.}
  \end{cases}
\end{equation}
The discretization policy discretizes $\cX$ into $k$ intervals.
Let the samples of $\cX$ in a given dataset $D$ be sorted in ascending order, ${\cx_{1:n} = \curly{\cx_1 \leq \cx_2 \leq \ldots \leq \cx_n}}$, and let the unique values be ${\cu_{1:m} = \curly{\cu_1 < \cu_2 < \ldots < \cu_{m}}}$.
The index of the last occurrence of $\cu_i$ in $\cx_{1:n}$ is denoted $s_i$.

@@

\begin{figure}[ht]
  \centering
  \input{Fig1}
  \caption{Factorization of $P(_{-\cX} \mid \Lambda)$.}
  \label{fig:example_factorization}
\end{figure}

@@

Some discretization edges appear in all three discretization methods, such as \num{1.42} and \num{2.35} for variable $C$ and \num{0.785} for variable $L$.
This indicates sufficiently MDL indeed can find some important discretization edges, but it is not sensitive to find more edges.

@@

\begin{table}[ht]
	\centering
	\caption{
		Discretization results for several datasets.
    The number of samples, number of variables, network structure, and mean cross-validated log-likelihood under each discretization method are given.
    The Bayesian method consistently has higher likelihood.
	}
  \resizebox{\textwidth}{!}{
	 \input{Table3_2.tex}
  }
	\label{table:more_data_set}
\end{table}

>>>>

Learning Bayesian networks from raw data can help provide insights into the relationships between variables. 
While real data often contains a mixture of discrete and continuous-valued variables, many Bayesian network structure learning algorithms assume all random variables are discrete.
Thus, continuous variables are often discretized when learning a Bayesian network.
However, the choice of discretization policy has significant impact on the accuracy, speed, and interpretability of the resulting models.
This paper introduces a principled Bayesian discretization method for continuous variables in Bayesian networks with quadratic complexity instead of the cubic complexity of other standard techniques.
Empirical demonstrations show that the proposed method is superior to the established minimum description length algorithm.
In addition, this paper shows how to incorporate existing methods into the structure learning process to discretize all continuous variables and simultaneously learn Bayesian network structures.

@@

Bayesian networks \citep{Pearl_1988,PGM_2009} are often used to model uncertainty and causality, with applications ranging from decision-making systems \citep{kochenderfer2012next} to medical diagnosis \citep{Lustgarten_2011}. Learning Bayesian networks from raw data can help users discover knowledge without field-specific experts, such as causality and probabilistic relation between variables. It is common to assume that the random variables in a Bayesian network are discrete, since many Bayesian network learning algorithms are unable to efficiently handle continuous variables. In addition, many of the commonly used Bayesian network software packages, such as Netica \citep{netica1992}, SMILearn \citep{druzdzel1999smile}, and bnlearn \citep{bnlearn2010}, are designed for discrete variables. However, many applications require the use of continuous data, such as positions and velocities in dynamic systems \citep{kochenderfer2010airspace}.

@@

Let $\cX$ be a continuous variable and let $\cx$ be a specific instance of $\cX$.
A discretization policy $\Lambda_{\cX} = \anglebrackets{e_1 < e_2 < \ldots < e_{k-1}}$ for $\cX$ is a mapping from $\mathbb{R}$ to $\curly{1,2,\ldots,k}$ such that
\begin{equation}
  \Lambda_{\cX}(\cx) = \begin{cases}
    1, & \text{if $x<e_1$}\\
    i, & \text{if $e_{i-1} \leq x < e_i$}\\
    k, & \text{otherwise.}
  \end{cases}
\end{equation}
The discretization policy discretizes $\cX$ into $k$ intervals.
Let the samples of $\cX$ in a given dataset $D$ be sorted in ascending order, ${\cx_{1:n} = \curly{\cx_1 \leq \cx_2 \leq \ldots \leq \cx_n}}$, and let the unique values be ${\cu_{1:m} = \curly{\cu_1 < \cu_2 < \ldots < \cu_{m}}}$.
The index of the last occurrence of $\cu_i$ in $\cx_{1:n}$ is denoted $s_i$.

@@

\begin{figure}[ht]
  \centering
  \input{Fig1}
  \caption{Factorization of $P(D_{-\cX} \mid \Lambda)$.}
  \label{fig:example_factorization}
\end{figure}

@@

Some discretization edges appear in all three discretization methods, such as \num{1.42} and \num{2.35} for variable $C$ and \num{0.785} for variable $L$. MDL can find some important discretization edges, but it is not sensitive to find more edges.

@@

\begin{table}[ht]
	\centering
	\caption{
		Discretization results for several datasets, summarizing the number of samples, number of variables, network structure, and mean cross-validated log-likelihood under each discretization method.
    The Bayesian method consistently has higher likelihood.
	}
  \resizebox{\textwidth}{!}{
	 \input{Table3_2.tex}
  }
	\label{table:more_data_set}
\end{table}
